---
title: "Collider bias for binary variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Collider bias for binary variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(AscRtain)
```

Following https://osf.io/vrcuf/, can infer the biased OR for a binary exposure ($A$) on a binary outcome ($Y$) when both of the traits influence the probability of being present in the sample ($S$)

Assume that being present in the sample is simply:

$$
P(S = 1 | A,Y) = \beta_0 + \beta_A A + \beta_Y Y + \beta_{AY} AY
$$

where $A = {0,1}$ and $Y = {0,1}$. The expected odds ratio under this scenario is then:

$$
E[\hat{OR}_{S=1}] = \frac{\beta_0(\beta_0 + \beta_A + \beta_Y + \beta_{AY})}{(\beta_0 + \beta_A)(\beta_0 + \beta_Y)}
$$

Suppose that we know the fraction of the population that is present in our sample ($p_{S}$). We are only interested in the $\beta_*$ parameter values that give rise to a value of $p_{S}$ that is within the bounds of expectation:

$$
p_{S} = \beta_0 + \beta_A p_A + \beta_Y p_Y + \beta_{AY} p_{AY}
$$

Scenario: we find an association between $A$ and $Y$ in our ascertained sample. Our question is what effects must $A$ and $Y$ have on sample ascertainment in order to induce the observed odds ratio $OR$, assuming that the true odds ratio is 1.

Initialise a new `VBB` (**V**-structure, **B**inary exposure, **B**inary outcome) class

```{r}
x <- VBB$new()
x
```

Example of how to calculate the odds ratio for given $\beta_*$ parameters:

```{r}
x$or_calc(b0=0.1, ba=0.2, by=0.3, bay=0.4)
```

Search over a parameter space of possible values to identify whether some target odds ratio could be explained by sample ascertainment

```{r}
x$parameter_space(
	target_or=2, 
	pS=0.0275, 
	pA=0.15,
	pY=0.1,
	pAY=0,
	b0_range=c(0,1), 
	ba_range=c(-0.2,0.2), 
	by_range=c(-0.2,0.2), 
	bay_range=c(0,0), 
	granularity=100
)
```

The parameter values that meet the target OR due to collider bias

```{r}
x$param
```

Visualise the distribution of odds ratios found across the range of parameters

```{r}
x$histogram()
```

Visualise the $\beta_*$ parameter ranges that meet the target odds ratio

```{r}
x$scatter()
```

Or in 3D:

```{r}
x$scatter3d()
```

Can try to do this in 3D also:

```{r}
plot3Drgl::plotrgl()
```


## Check the results through simulation

Here, $A$ is ACE inhibitor use, $Y$ is Covid-19 status, and $S$ is presence in the first release of the COVID Symptom Tracker dataset.

Observational association of ACE-i influence on Covid-19 status gives OR $\approx 2$. Assume 15% of population take ACE-i, 10% are infected with coronavirus at the time of sampling, 1.9 million of an adult population of 54 million are present in the sample (3.5%). What influences of ACE-inhibitor use and Covid-19 status would be required to induce a collider bias of $OR=2$?

Enter parameters:

```{r}
x <- VBB$new()
x$parameter_space(
	target_or=2, 
	pS=0.035, 
	pA=0.15,
	pY=0.1,
	pAY=0,
	b0_range=c(0,1), 
	ba_range=c(-0.2,0.2), 
	by_range=c(-0.2,0.2), 
	bay_range=c(0,0), 
	granularity=100
)
x$scatter3d()
```

Simulate individual level data according to the first result, where $A$ has no influence on $Y$. Note that using a population size of 1 million as the sample size doesn't matter, just the proportion sampled.

```{r}
a <- subset(x$param, abs(by) < 0.1 & b0 < 0.02)[1,]
n <- 1000000
Y <- rbinom(n, 1, a$pY)
A <- rbinom(n, 1, a$pA)
pS <- a$b0 + a$ba * A + a$by * Y
S <- rbinom(n, 1, pS)
```

What proportion of the population are present in the sample?

```{r}
sum(S) / length(S)
```

Estimate association between `A` and `Y`

```{r}
summary(glm(Y[as.logical(S)] ~ A[as.logical(S)], family="binomial"))$coef[2,1] %>% exp
```

